---
title: "lab_02"
author: "Adelia McGuire"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Helpful Functions

```{r}
head(md_voters_2024)
summary(md_voters_2024)
colnames(md_voters_2024)
glimpse(md_voters_2024)
```

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1** Montgomery County has seen the most amount of new registrants since July 21st with a total of 4847. This is followed by Baltimore City with 2799 and Baltimore County with 1683. When we are comparing which counties have seen the most amount of new registrants since July 21 we should first look at population. Comparing Montgomery's 4847 with Somerset's 31 population plays a key role in the stark difference in new registrants. As of 2022 Somerset had a population of 346,875 while Montgomery had a population of 1.053 million the same year.

```{r}
new_voters <- 
md_voters_2024 |>
 
  filter(STATE_REGISTRATION_DATE==COUNTY_REGISTRATION_DATE) |>
    
  filter(STATE_REGISTRATION_DATE>'2024-07-21') 
  
  head(new_voters)
```

```{r}
  new_voters |>
  group_by(COUNTY) |>
  summarise(
    county_totals = n()
  ) |>
  arrange(desc(county_totals))
```

**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** Create a new data frame allowed me to clearly see the total number of voters per week. In week one there were 2029 total voters, 2085 in week two, 2003 in week three, 3223 in week four, 3524 in week five, and 3395 in week six. In general I noticed that the number of voters per week increased as the weeks progressed. In the first three weeks we saw a voter average of just 2039. In the last three weeks the voter average was around 3380. 

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(week_number = week(COUNTY_REGISTRATION_DATE))

voters_per_week <- md_voters_2024 |>
  group_by(week_number) |>
    summarise(weekly_totals = n()) |>
    arrange(week_number)
  
head(voters_per_week)
```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** Through using the lubridate month() we are able to gain access to the month from COUNTY_REGISTRATION_DATE. The function then works to group by each different month and to count the total number of registrations for each individual month.As seen below, I am ensuring that the name of each month is displayed properly by using "label = TRUE". These findings tell us that the month of July had the highest monthly voter counts with a total of 26284. 

```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(month = month(COUNTY_REGISTRATION_DATE, label = TRUE))

monthly_voter_counts <- md_voters_2024 |>
  group_by(month) |>
  summarise(monthly_totals = n()) |>
  arrange(desc(monthly_totals))

monthly_voter_counts

```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** When going about this code I felt like I had two potential routes, I could use ifelse or case_when. I felt that modifying the original dataframe using case_when was a clear and easy to follow way of demonstrating we are implementing a new column (called "region"). Although listing all of the individual names of the counties was rather tedious, it was ultimately effective in showing what regions and parties account for the most new voters. According to my data, the region that accounts for the most amount of new voters is the Capitol Region with a whopping 14,837 people. The Democratic party also accounts for the most new voters. 

```{r}
md_voters_2024 <- md_voters_2024 |>
  
  mutate(region = case_when(
    COUNTY %in% c('Montgomery', 'Frederick', 'Howard') ~ 'Capital Region',
    COUNTY %in% c('Baltimore City', 'Baltimore County', 'Harford') ~ 'Central Maryland',
    COUNTY %in% c('Carroll', 'Washington', 'Allegany', 'Garrett') ~ 'Western Maryland',
    COUNTY %in% c('Calvert', 'Charles', 'Prince George\'s', 'St. Mary\'s') ~ 'Southern Maryland',
    COUNTY %in% c('Cecil', 'Dorchester', 'Kent', 'Queen Anne\'s', 'Somerset', 'Talbot', 'Wicomico', 'Worcester') ~ 'East Shore',
    TRUE ~ 'Other'
  ))

region_party <- md_voters_2024 |>
  group_by(region, PARTY) |>
  summarise(region_totals = n()) |>
  arrange(desc(region_totals))


head(region_party)

```

**Q5** What's an additional question you could ask of this data?

**A5** When analyzing data it is important to keep in mind the questions you still have and what you would benefit from knowing. I think that asking questions surrounding age and new voter registration. One solid question could be: How does new voter registration vary depending on age group. Another could be: Which age group has the most politically active new voters? 
