---
title: "pre_lab_04.Rmd"
author: "Adelia McGuire"
date: "2024-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructor demonstration points to hit

1.  Ensure students install refinr package and load it.
2.  Show students how to download and install Open Refine and confirm install before they leave class.

## Chapter 16

### Task 1: Install refinr package

**Task** Run the following code in the gray-colored codeblock below to install refinr.

```{r}
install.packages('refinr')
```

### Task 2: Load libraries and settings

**Task** Run the following code in the gray-colored codeblock below to load the tidyverse, janitor, and refinr and turn off scientific notation.

```{r}
# turn off sci notation
options(scipen=999)
library(tidyverse)
library(lubridate)
library(refinr)
```

### Task 3: Load data

**Task** Run the following code to load some campaign expenditure data focused on food-related expenses in Washington, D.C. 

**Answer** The following code provides us with campaign expenditure data that specifically pertains to food related expenses in Washington, D.C.  

```{r}
# Load data
dc_food <- read_csv("data/dc_food.csv")

# Display it
dc_food

```

### Task 4: Group and count to identify problems in city column

**Task** Run the following code to group and count records in the recipient_name column, and filter to examine only recipients that start with W. Name a problem that you see that is preventing proper grouping. 

**Answer** One issue that I see in this data is that there are a multitude of duplicates in the recipient name column. This error of duplication is due to a variation in spelling that confuses the data, an example can be seen with W Millar & Co. Catering which is listed under, "W Millar Catering", "W. MILLAR & CO, W. MILLAR & CO.", "W. MILLAR & CO. CATERING", "W. Millar & Co.", and "W.Millar & Company Catering". Each name varies in spelling/formatting but is all referring to the same dining location. This problem occurs for multiple recipient names within the data. Another issue is that "filter(str_detect(recipient_name, '^w|^W'))" is separating each sorting the data based on if it is all caps or lowercase, failing to consider that their are duplicates of names that are both lowercase and uppercase. An example of this situation is with the recipient Walmart which is listed first in uppercase as "WALMART" and then later in lowercase as "Walmart". 

```{r}
# Now let's try and group and count the number of expenditures by recipient. To make it a bit more manageable, let's use another string function from `stringr` and filter for recipients that start with the uppercase "W" or lowercase "w" using the function `str_detect()` with a regular expression.  

# The filter function in the codeblock below says: look in the city column, and pluck out any value that starts with (the "^" symbol means "starts with") a lowercase "w" OR (the vertical "|", called a pipe, means OR) an uppercase "W". 

dc_food |>
  group_by(recipient_name) |>
  summarise(
    count=n()
  ) |>
  filter(str_detect(recipient_name, '^w|^W')) |> 
  arrange(recipient_name)
```

### Task 5: Use refinr to clean the city_clean field.

**Task** Run the following code to make a copy of the recipient_name field called recipient_clean, then use the `key_collision_merge` function from refinr to attempt to standardize similar values. Store it as a new object called cleaned_dc_food. Click the black arrow on the table to examine the recipient_name and recipient_clean columns. Looking at the results on the first two pages only (first 20 rows), name two changes the refinr package made, by comparing recipient_name to recipient_clean. 

**Answer** The key_collision_merge function from the refinr package changed the column recipient_clean by taking standardized variations in the recipient_name field and merging similar entries into one consistent wording format. Where their were once different spellings and representations of the same dining entitity, such as "W Millar & Co. Catering" and "W. MILLAR & CO.," the titles were consolidated into "W Millar & Co." Another change that came from the key_collision_merge function was the removal of extraneous punctuation and unnecessary words. Each of these changes were important in their ability to enhance general data accuracy and an overall more reliable analysis, ensuring that unintentional duplicates are only represented as a single entity.

```{r}
cleaned_dc_food <- dc_food |>
  mutate(recipient_clean=key_collision_merge(recipient_name)) |>
  select(recipient_name, recipient_clean, everything())

cleaned_dc_food

```

### Task 6: Filter cleaned_dc_food to examine recipients that start with W

**Task** Run the following code to group by recipient_clean and recipient_name, then filter for cities that start with W. Name three changes the refinr function made in the first 20 results. 

**Answer** The refinr package has multiple ways to clean up and resolve inconsistencies. In task 6 the function the refinr significantly improved the organization and clarity of the recipient_name data by implementing three key changes. First, it consolidated variations of the same recipient into a single standardized name to create consistency in declaring a single consistent entry. This first factor, of consolidation one which I have spoken on throughout this prelab in the W Millar & Co. example., reduces redundancy of the dataset. The second way the function alter sthe presentation of the data is in its removing of confusing extraneous characters, such as certain punctuation marks or extra spaces, changing names for improved readability. Lastly, by introducing the recipient_clean column, the analysis was refined to focus strictly on standardized names, increasing the accuracy in  filtering for entries starting with "W" or "w." 

```{r}
cleaned_dc_food |>
  group_by(recipient_name, recipient_clean) |>
  summarise(
    count=n()
  ) |>
  filter(str_detect(recipient_clean, "^w|^W")) |>
  arrange(recipient_clean)
```

### Task 7: Use a different refinr method to clean the city_clean field.

**Task** Run the following code to use the n_gram_merge() function to clean the data, then group by city_clean and city to examine cities that start with A. Look at the differences. Does the changes it made vary substantially from the previous use of key_collision_merge?

**Answer** In comparing this code to the former code it is clear that the n_gram_merge() function from the refinr package provides an different unique approach to cleaning the recipient_name field. While key_collision_merge focuses on merging exact matches and handling common variations, n_gram_merge() is more analytical, looking at the names based on substrings and common patterns. This means that  n_gram_merge()can quickly identify/merge entries that may not be exact matches but still share very significant components. This ultimately makes for an enhanced data cleaning process. This can be seen for instance in how names like "W. Millar & Co." and "W Millar Catering" may be better consolidated through this method, regardless of if they contain different formatting. I feel that the changes made by n_gram_merge() often produce a larger consolidation of entries, honing in on and capturing variations that key_collision_merge otherwise may have overlooked. This means that the n_gram_merge() likely will produce fewer unique names in the recipient_clean column, ultimately working to improving data accuracy and reduce redundant recipient names. In my experience anyalzing these two different methods used to standardize data, n_gram_merge() boasts a more nuanced approach to data standardization that can be particularly useful in situations that call for working with many similar entries. 


```{r}
cleaned_dc_food <- dc_food |>
  mutate(recipient_clean=n_gram_merge(recipient_name)) |>
  select(recipient_name, recipient_clean, everything())

cleaned_dc_food


cleaned_dc_food |>
  group_by(recipient_name, recipient_clean) |>
  summarise(
    count=n()
  ) |>
  filter(str_detect(recipient_clean, "^w|^W")) |>
  arrange(recipient_clean)


```

### Task 8: Follow the directions in the textbook to use Open Refine

**Task** Follow the directions in the textbook to upload dc_food.csv to Open Refine (<https://thescoop.org/data_journalism_book/open-refine.html#manually-cleaning-data-with-open-refine>). Follow the instructions to clean as many recipients as the Open Refine clustering tool can identify. In class, I will come around to look at the Open Refine project you created to do this.

**Answer** Completed! 
