---
title: "lab_08"
author: "Adelia McGuire"
date: "2024-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lab we'll be looking at Maryland absentee ballot requests in the 2024 election and making several charts based on that data. The state of Maryland reports the number of ballots sent to voters by counties and the number received by counties from voters. We'll use both types of information.

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts and cleaning up data.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load libraries
library(tidyverse)
library(ggthemes)
library(janitor)
library(tidyr)
```

## Load data

**Task** Create a codeblock and load the following data from the data folder:

-   Maryland absentee ballots sent and returned - `md_absentee_ballots.csv`
-   Maryland absentee ballots sent and returned by date - `md_absentee_ballots_date.csv`

You can clean up column names to make them easier to work with if you like, and should ensure that any date columns are actually date formatted.

```{r}
# load the data
md_absentee_ballots <- read_csv("~/Downloads/GitHub - JOUR472/GitHub Folder/data_journalism_2024_fall/labs/lab_08/data/md_absentee_ballots.csv")
md_absentee_ballots_dates <- read_csv("~/Downloads/GitHub - JOUR472/GitHub Folder/data_journalism_2024_fall/labs/lab_08/data/md_absentee_ballots_date.csv")
```

## Questions

**Q1.** Which county has the highest percentage of total absentee ballots returned of total ballots sent? Make a bar chart of the top 10 counties. Your bar chart must have:

-   A clear title that states the main idea/finding
-   Good labels for the x & y axis and a caption for the source, which is the Maryland State Board of Elections
-   Readable bars - the values shouldn't be overlapping

**A1.** The Maryland county with the highest percentage of total absentee ballots returned of the total ballots sent is Worcester County with 62.59429%. This ranking that shows Worcester in the top stop of the 10 Maryland counties with the highest percentage is reflected both in the " highest_absentee_ballot_percentage" datablock and bar chart. 

```{r}
md_absentee_ballots <- md_absentee_ballots |>
  clean_names()|>
  mutate(absentee_percentage = (total_received / total_sent) * 100)

highest_absentee_ballot_percentage <- md_absentee_ballots |>
  arrange(desc(absentee_percentage)) |>
  select(county_name, absentee_percentage) |> 
  head(10) 

highest_absentee_ballot_percentage
```

```{r}
highest_absentee_ballot_percentage |>
  ggplot() +
  geom_bar(aes(x=reorder(county_name,absentee_percentage), weight=absentee_percentage)) +
  coord_flip() + 
  theme_economist() +
  labs(
    title="Counties with Highest Absentee Ballot Percentage",
    x = "County Name",
    y = "Absontee Ballot Percentage",
    caption = "Source: Maryland State Board of Elections"
  )
```
**Q2.** What are the top 10 counties with the lowest percentage of ballots returned by Democratic voters, and what is the difference between the percentage returned for Democrats and Republicans in those counties? You MUST do the following things when answering this question:

-   Make a codeblock below to write your code in.
-   Calculate the percentage of ballots returned by both Democrats and Republicans.
-   Arrange the dataframe to show the counties with the smallest percentage of returned ballots from Democrats first.
-   Use ggplot to make a horizontal bar chart that shows just the first 10 counties, with the length of the bar determined by the county's percentage of ballots returned by Democrats. Give it an appropriate title, source, and x and y axis titles.

**A2.** The top 10 counties with the lowest percentage of ballots returned by Democratic voters, beginning with the lowest percentage, are... Baltimore City, Montgomery County, Prince George's County, Frederick County, Baltimore County, Charles County, Howard County, Cecil County, Anne Arundel County, and Carroll County. The top 10 counties with the lowest percentage of ballots returned by Republican voters, beginning with the lowest percentage, are... Baltimore City, Montgomery County, Prince George's County, Frederick County, Howard County, Baltimore County, Cecil County, Saint Mary's County, Calvert County, and Carroll County. Both the county with the lowest percentage of ballots returned by Democratic voters(40.47761%) and the county with the lowest percentage of ballots returned by Republican voters(34.61140) is Baltimore City. When comparing the similarities and differences between the lowest percentage of ballots returned by Republicans and Democrats it is clear that the counties that have higher populations their is generally a lower ballot return percentage. 

```{r}
dem_absentee_ballots <- md_absentee_ballots |>
  clean_names()|>
  mutate(democratic_absentee_percentage = (dem_received / dem_sent) * 100)

lowest_democratic_absentee_ballot_percentage <- dem_absentee_ballots |>
  arrange(democratic_absentee_percentage) |>
  select(county_name, democratic_absentee_percentage) |> 
  head(10) 

lowest_democratic_absentee_ballot_percentage
```

```{r}
lowest_democratic_absentee_ballot_percentage |>
  ggplot() +
  geom_bar(aes(x=reorder(county_name,-democratic_absentee_percentage), weight=democratic_absentee_percentage)) +
  coord_flip() + 
  theme_economist() +
  labs(
    title="Lowest Dem Absentee Ballot Percentage County",
    x = "County Name",
    y = "Democratic Absontee Ballot Percentage",
    caption = "Source: Maryland State Board of Elections"
  )
```

```{r}
rep_absentee_ballots <- md_absentee_ballots |>
  clean_names()|>
  mutate(republican_absentee_percentage = (rep_received / rep_sent) * 100)

lowest_republican_absentee_ballot_percentage <- rep_absentee_ballots |>
  arrange(republican_absentee_percentage) |>
  select(county_name, republican_absentee_percentage) |> 
  head(10) 

lowest_republican_absentee_ballot_percentage
```

**Q3** Let's make a line chart using the second set of absentee ballot data - this one tracking returns by date. We want to see received ballots for Frederick County for Democrats, Republicans and Others, so you'll make a chart with three lines, adding color to each one (you can choose, but choose sensibly). Give the chart a title and good labels. What's the story that this line chart tells? Compare and contrast the three lines. Does it match what you know about absentee voting habits?

**A3** This line chart that I have created works to visually show the trends in absentee ballots received in Frederick County. The chart breaks down the data by pary affiliations, showing the trends over time of Democrats, Republicans, and Other. The blue line represents the ballots received from Democrats and it shows a consistent upward trend as time passes and the election date approaches. This could serve to indicate that the democratic party is making proactive mobilization efforts or additionally that their is for some reason a higher chance of Democratic voters using absentee ballots. The red line for Republicans displays a more all over the place pattern, with fairly frequent fluctuations in ballot returns. While it does show that their is a slight peak as the election date nears, this peak is not as severe to that steady increase we observed in the Democratic line. The purple line represents the 'Others' category that covers voters who are not affiliated with the Democratic or Republican parties. This line is fairly low compared to the other two lines though it  may indicate a slight increase as the election date nears. I find that this line chart tells a story about the differences in engagement among the parties by comparing the absentee voting trends in Frederick County. The noticeable increase for Democrats as we get closer to November 5th contrasts with the Republicans more variable pattern. This data is very useful in identifying members of the parties absentee voting habits. 

```{r}
md_absentee_ballots_dates <- md_absentee_ballots_dates |> 
  clean_names() 

md_absentee_ballots_by_date <- md_absentee_ballots_dates |>
  filter(county_name == "FREDERICK") |>
  group_by(date) |>
  summarise(
    dem_received = sum(dem_received, na.rm = TRUE),
    rep_received = sum(rep_received, na.rm = TRUE),
    oth_received = sum(oth_received, na.rm = TRUE),  
    .groups = 'drop'
  )

md_absentee_long <- md_absentee_ballots_by_date |>
  pivot_longer(cols = c(dem_received, rep_received, oth_received), 
               names_to = "party", 
               values_to = "ballots_received")

ggplot(md_absentee_long, aes(x = date, y = ballots_received, color = party)) +
  geom_line(size = 1) +
  labs(title = "Absentee Ballots Received by Party in Frederick County",
       x = "Date",
       y = "Number of Ballots Received") +
  scale_color_manual(values = c("blue", "purple", "red")) +
  theme_minimal() +
  theme(legend.title = element_blank())
```

**Q4.** A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis. For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid.

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable. A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot to get a visual sense of two key variables: ballots sent and ballots returned. We want to plot both of those for each county in Maryland. IMPORTANT: you need to choose whether you use total ballots or one of the parties.

Follow these steps:

1. Using the first dataframe (the one without dates), make a scatterplot showing the number of ballots sent on one axis and the number received on the other axis. The pre_lab didn't show you how to do this, so look it up! https://ggplot2.tidyverse.org/reference/geom_point.html?q=geom_point#aesthetics is a good start.
2. Give it an appropriate title, source, and x and y axis titles.
3. Try to add a label for each point that shows the name of the county using geom_text(). You might need to find alternative ways to do this to make them readable.
4. Add a line of fit to the plot using geom_smooth (https://ggplot2.tidyverse.org/reference/geom_smooth.html#ref-examples). Specify the method as lm.
5. In the answer space below, describe what you see. In a general sense, how would you report out the main point(s) of this chart? Are there counties that seem to be over or under-performing in relation to the overall trend?

**A4.** I created a scatterplot that reflects the number of ballots sent on the X axis and the number of ballots received or returned on the Y axis. To make sure you can easily identify the location of the different counties on the line I made the points green and the line itself navy. Through observing my scatterplot I notice that Montgomery County seems to be something of an outlier in the data as it has the highest number for both the total ballots received and the total ballots sent. In relation to the overall trend it appears that Baltimore County, Prince George's County, and Anne Arundel County are all over performing when it comes to their number of total ballots returned. I would report out the main points of this chart by bringing attention to these counties that I have talked about that are over achieving (and/or underachieving) or are considered outliers in the general data. 

```{r}
md_absentee_ballots <- md_absentee_ballots|> 
  clean_names() 

ggplot(md_absentee_ballots, aes(x = total_sent, y = total_received, label = county_name)) +
  geom_point(color = "green", size = 3) +  
  geom_text(vjust = -1, size = 2.5) +  
  geom_smooth(method = "lm", color = "navy", se = FALSE) +
  labs(title = "Relationship Between Maryland Ballots Sent and Returned",
       x = "Number of Total Ballots Sent",
       y = "Number of Total Ballots Returned",
       caption = "Source: Maryland Board of Elections") +
  theme_minimal()
```
