---
title: "pre_lab_08.Rmd"
author: "Adelia McGuire"
date: "2024-10-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 21

Visualizing data is becoming a much greater part of journalism. Large news organizations are creating graphics desks that create complex visuals with data to inform the public about important events.

To do it well is a course on its own. And not every story needs a feat of programming and art. Sometimes, you can help yourself and your story by just creating a quick chart, which helps you see patterns in the data that wouldn't otherwise surface.

Good news: one of the best libraries for visualizing data is in the tidyverse and it's pretty simple to make simple charts quickly with just a little bit of code. It's called [ggplot2](https://ggplot2.tidyverse.org/).

Let's revisit some data we've used in the past and turn it into charts. First, let's load libraries. When we load the tidyverse, we get ggplot2.

### Task 1: Load packages

**Task** Run the following code to load packages.

```{r}
library(tidyverse)
```

The dataset we'll use is voter registration data by county in Maryland from 2020 and September 2022. Let's load it.

### Task 2: Load data

**Task** Run the following code to load the data. In the space below, indicate how many rows and columns it has. 

**Answer** The following code, which is titled "maryland_voters_2024_2022", has a total of 24 rows and 18 columns. 

```{r}
md_voters <- read_csv("data/maryland_voters_2024_2022.csv")
```

## Bar charts

The first kind of chart we'll create is a simple bar chart.

It's a chart designed to show differences between things -- the magnitude of one thing, compared to the next thing, and the next, and the next.

So if we have thing, like a county, or a state, or a group name, and then a count of that group, we can make a bar chart.

So what does the chart of the top 10 maryland counties with the biggest change in unaffiliated voters from 2024 to 2022 look like?

First, we'll create a dataframe of those top 10, called maryland_top_counties.

### Task 3: Create a dataframe of top 10 counties for unaffiliated voter change in Maryland

**Task** Complete and run the following code to create a dataframe of top 10 counties with the biggest change in unaffiliated voters in Maryland. In the space below, explain what each step of the code below is doing in English. 

**Answer** In order to create an effective and efficient code it is critical one first knows the exact function and purpose of each step of their code. The following code creates a dataframe that shows the top 10 countries with the biggest change in unafilliated voters in Maryland, to understand how this dataframe was obtained I will break down each piece of this code. The first step can be found in the lines md_voters <- md_voters |> mutate(UNAF_DIFF = UNAF_2024 - UNAF_2022) and is responsible for creating a new dataframe for difference in unaffiliated voters. This line or lines takes the existing dataframe md_voters and adds a new column to it called UNAF_DIFF. This new column works to calculate the difference between the number of unaffiliated  voters in 2024 ad the number of unaffiliated voters in 2022. Additionally, the mutate function is what creates this new column in the dataframe. The next step, represented in lines "maryland_top_counties <- md_voters |> arrange(desc(UNAF_DIFF)) |> select(County, UNAF_DIFF) |>", is how the code arranges counties by the amount of change in unaffiliated voters.  The code takes the updated md_voters dataframe and first arranges is by sorting the data in descending order so that the counties with the greatest increase in unaffiliated voters will show first before using the select function to select only the "County" and "UNAF_DIFF" Finally, the last step is responsible for displaying and presenting the data. The lines, "head(10) maryland_top_counties", retrieves the top 10 rows from the sorted dataframe to show us the 10 counties with the largest change in the number of unaffiliated voters and then also outputs the maryland_top_counties dataframe. 
```{r}
md_voters <- md_voters |> 
  mutate(UNAF_DIFF = UNAF_2024 - UNAF_2022)

maryland_top_counties <- md_voters |>
  arrange(desc(UNAF_DIFF)) |>
  select(County, UNAF_DIFF) |> 
  head(10) 

maryland_top_counties
```

Now let's create a bar chart using ggplot.

With ggplot, the first thing we'll always do is draw a blank canvas that will contain our chart. We start with our dataframe name, and then (%\>%) we invoke the ggplot() function to make that blank canvas. All this does is make a gray box, the blank canvas that will hold our chart.

### Task 4: Create an empty ggplot container

**Task** Run the following code. Explain what you see after running the code in the space below. 

**Answer** When I run this code I see a blank grey box or canvas. Nothing is displayed, it is an empty plot area. 

```{r}
maryland_top_counties |>
  ggplot()
```

Next we need to tell ggplot what kind of chart to make.

In ggplot, we work with two key concepts called geometries (abbreviated frequently as geom) and aesthetics (abbreviated as aes).

Geometries are the shape that the data will take; think of line charts, bar charts, scatterplots, histograms, pie charts and other common graphics forms.

Aesthetics help ggplot know what component of our data to visualize -- why we'll visualize values from one column instead of another.

In a bar chart, we first pass in the data to the geometry, then set the aesthetic.

In the codeblock below, we've added a new function, geom_bar().

Using geom_bar() -- as opposed to geom_line() -- says we're making a bar chart.

Inside of that function, the aesthetic, aes, says which columns to use in drawing the chart.

We're setting the values on the x axis (horizontal) to be the name of the county. We set weight to total loans, and it uses that value to "weight" or set the height of each bar.

One quirk here with ggplot.

After we've invoked the ggplot() function, you'll notice we're using a + symbol. It means the same thing as %\>% -- "and then do this". It's just a quirk of ggplot() that after you invoke the ggplot() function, you use + instead of %\>%. It makes no sense to me either, just something to live with.

### Task 5: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. 

**Answer** When I run the following code I see a bar chart with a plot that displays bars that correspond to each county that is listed in the maryland_top_counties dataframe. The height of each bar is dependent upon the change in the number of unaffiliated voters, meaning that the taller bars show a higher increase in unaffiliated voters between 2022 and 2024 while the shorter bars show a smaller increase. Each of the counties are plotted on the x-axis and the height of each bar is reflective to how much that county's unaffiliated voter count has changed over the 2 year time span. The chart is visually showing which counties experienced the most notable shifts (increases or decreases) in voter affiliation.

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=County, weight=UNAF_DIFF))
```

This is a very basic chart. But it's hard to derive much meaning from this chart, because the counties aren't ordered from highest to lowest by total_loans. We can fix that by using the reorder() function to do just that:

### Task 6: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? 

**Answer** Upon running the following code, I see a bar chart with a plot that shows Maryland counties organized from the least to the most change in unaffiliated voters.This ordering makes it much easier to compare just how much change each county saw and provides the viewer with a clear and straightforward answer as to which counties saw the most significant shifts. The change from the last code and corresponding plot is that the counties are now ordered by the extent of change in unaffiliated voters, rather than just arranged by their original order from the dataset.

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF))
```

This is a little more useful. But the bottom is kind of a mess, with overlapping names. We can fix that by flipping it from a vertical bar chart (also called a column chart) to a horizontal one. coord_flip() does that for you.

### Task 7: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? 

**Answer** After running this code I see a horizontal bar chart that shows the change in unaffiliated voters for the top 10 counties in Maryland. This formerly vertical chart is transformed to a horizontal plot due to the addition of coord_flip(), changing the counties to display on the y-axis rather than the x-axis. The heights of the bars which represent the change in unaffiliated voters extend out horizontally rather than vertically as well. This plot displays horizontal bars for each county which does improve the readability of the data by making it easier for us to read the potentially long county names. One thing that has remained the same is that the height or length of the bars still reflects the extent of the change in unaffiliated voters. The most noticeable change in this code and how the data now presents itself is that the bars show up horizontal rather than their formerly vertical orientation. In general, I feel that this specific format allow the user to make better comparisons of the values because the x-axis is not jam packed with names that are closely spaced together and therefor hard to read.  

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip()
```

Is this art? No. Does it quickly tell you something meaningful? It does.

We're mainly going to use these charts to help us in reporting, so style isn't that important.

But it's worth mentioning that we can pretty up these charts for publication, if we wanted to, with some more code. To style the chart, we can change or even modify the "theme", a kind of skin that makes the chart look better.

It's kind of like applying CSS to html. Here I'm changing the theme slightly to remove the gray background with one of ggplot's built in themes, theme_minimal().

### Task 8: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? 

**Answer** When running this updated code a horixontal bar chart is revealed. This horizontal bar chart, though similar to the one in the previous code block, is significantly different in how it is visually styled because of the incorporation of the theme_minimal() function. This function works to apply a clear, minimalist theme to the plot, implementing changes such as a simplified background. In terms of changes, although the plot shows the same horizontal bars for each county, the overall appearance is cleaner to bring focus strictly to the data. This application of a minimalist theme is the primary change due to how the general visual appeal and clarity of the plot is enhanced. 
```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip() + 
  theme_minimal()
```

The ggplot universe is pretty big, and lots of people have made and released cool themes for you to use. Want to make your graphics look kind of like [The Economist's](https://www.economist.com/) graphics? There's a theme for that.

First, you have to install and load a package that contains lots of extra themes, called [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

### Task 9: Run the following code

**Task** Run the following code. 

**Answer** The following code was succesfully run and ggthemes was downloaded... "The downloaded binary packages are in
	/var/folders/v_/5cgv7gdn3lv9fc_h1jy4bv940000gn/T//RtmpzytfaU/downloaded_packages" 

```{r}
install.packages('ggthemes')
library(ggthemes)
```

And now we'll apply the economist theme from that package with theme_economist()

### Task 10: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last version of the chart? 

**Answer** The following code reveals a horizontal bar chart that is similar to the previus one but very different visually due to the incorporation of the the function, "theme_ecobomist(), which is a clean and professional looking theme. The bars are ow dark grey, contrasting again the light blue background and white grid lines. The names of each county are also dark grey which makes them easy to read against the background as well. The primary change between this chart and the last chart is the application of the Economist theme and how it gives the plot a very neat and professional general vibe. Readability and visual components are enhanced even further

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip() + 
  theme_economist()
```

Those axis titles are kind of a mess. Let's change "count" on the x axis to "net change" and change "reorder(County,TOTAL_DIFF)" to "county". And while we're at it, let's add a basic title and a source as a caption. We'll use a new function, labs(), which is short for labels.

### Task 11: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? 

**Answer** The plot that I see from this code has a clear and bold title at the top, labeled axes, and source citation in the bottom right corner. Everything about this plot is easier to understand and very informative due to the visual components of the Economist theme. This code and plot have changed in that there is not a title, axis labels, and a caption, each of which works to enhance the chart. 

```{r}
maryland_top_counties |>
  ggplot() +
  geom_bar(aes(x=reorder(County,UNAF_DIFF), weight=UNAF_DIFF)) +
  coord_flip() + 
  theme_economist() +
  labs(
    title="Counties with Biggest Jumps in Unaffiliated Voters",
    x = "net change",
    y = "county",
    caption = "source: Maryland State Board of Elections"
    
  )
```

Viola. Not super pretty, but good enough to show an editor to help them understand the conclusions you reached with your data analysis.

## Line charts

Let's look at how to make another common chart type that will help you understand patterns in your data.

Line charts can show change over time. It works much the same as a bar chart, code wise, but instead of a weight, it uses a y.

So, let's load some WinRed contribution data we've previously used and create a dataframe with a count of contributions for each date in our data.

### Task 12: Load the WinRed contribution data and create a dataframe of contributions by date

**Task** Run the following code to load the WinRed data

```{r}
md_winred <- read_rds("data/maryland_winred.rds")
```

**Task** Run the following code. Explain what you see after running the code in the space below. 

**Answer** Upon running this code I see an output which is a new dataframe that displays each unique date from the md_winred dataset next to its corresponding amount of contributions for that specific date. This code works to take the original dataset and transform it into something of a simpler format that easily brings light to the total number of contributions received on each date. 


```{r}

md_winred_by_date <- md_winred |>
  group_by(date) |>
  summarise(
    total_contributions=n()
  )

md_winred_by_date 
```

And now let's make a line chart to look for patterns in this data.

We'll put the date on the x axis and total contributions on the y axis.

### Task 13: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. 

**Answer** The resulting plot shows a line graph where the x-axis shows the dates and the y-axis shows the total number of contributions recorded on those specific dates. I see a line that connects data points that represent the total number of contribution for each date, ultimately displaying a trend that takes place over time. This code and the line graph which it produces is useful in that it provides a visual representation of how contributions can vary over time, displaying quick increases/decreases on certain dates. 

```{r}
md_winred_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_contributions))


```

It's not super pretty, but there's a pattern here: the number of contributions fluctuates between 400 and 800 a day for most of this period, and then jumps way up at the end of June. We've learned that the end of June is the end of a reporting period, and donors respond to deadlines.

Right now, it's kind of hard to see specifics, though. When did some of those smaller spikes and troughs happen?

We can't really tell. So let's modify the x axis to have one tick mark and label per month. We can do that with a function called scale_x_date().

We'll set the date_breaks to appear for every week; if we wanted every month, we'd say date_breaks = "1 month". We can set the date to appear as month abbreviated name (%b) and day (%d).

### Task 14: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. 

**Answer** Upon running this code, I see a plot that displays a line graph where the x-axis shows dates at weekly intervals, ultimately boosting readability through reducing the number of ticks on the plot. Additionally, I took note to how the date labels are formatted in a way that clearly shows the month and day, making it easy to quickly identify the timeline of contributions.

```{r}
md_winred_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_contributions)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d")


```

Those are a little hard to read, so we can turn them 45 degrees to remove the overlap using the theme() function for styling. With "axis.text.x = element_text(angle = 45, hjust=1)" we're saying, turn the date labels 45 degrees.

### Task 15: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below.

**Answer** I see a plot displaying a line graph that has the x-axis showing dates at weekly intervals, with the labels rotated to 45 degrees. This unique orientation is really helpful when it comes to to clearly displaying the dates without them over-crowding one another. The primary change is the rotation of the x-axis labels, improving the plot by allowing for a far more easier interpretation of trends across long periods of time in the contributions data. 

```{r}
md_winred_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_contributions)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  theme(
    axis.text.x = element_text(angle = 45,  hjust=1)
  )

```

Again, this isn't as pretty as we could make it. But by charting this, we can quickly see a pattern that can help guide our reporting.

We're just scratching the surface of what ggplot can do, and chart types. There's so much more you can do, so many other chart types you can make. But the basics we've shown here will get you started.
