---
title: "pre_lab_01.Rmd"
author: "Adelia McGuire"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```

## About this notebook

This notebook contains code and explanatory text that your should review and run as you read through chapters 1 - 5 of the course textbook, "Data Journalism with R and the Tidyverse". Answer questions and edit the document as directed.

Running this notebook will help you understand key data analysis methods and concepts that you will put into practice during this week's lab. When you come across a new command or verb in R, you should add it to your class reference notebook file along with what it's used for.

When you are finished running the code in this notebook, you will push changes to your course GitHub repo, and upload the link to ELMS as instructed.

You also should update your class reference notebook with code examples and what they mean. When you come across a new concept or R verb, put it in your reference notebook along with a sentence about what it does. For example:

```{r}
library(tidyverse)
```

The library function loads an R library such as the tidyverse, providing access to the verbs and functions that library contains. You have to load a library before you can use its functions.

## Chapter 3

### Task 1: The console

**Task** In the console (look below), type 2+2 and hit enter. What does it return? Type your answer below. **Answer:**
Answer: 4

### Task 2: Variables

**Task** Put the following code -- which stores the numeral 2 as a variable called number -- in the console and hit enter.

```{r}
number <- 2
```

Then type this code to multiply the variable called number by itself.

```{r}
number * number
```

What does it return? Type your answer below:

**Answer:**
Answer: 2 * 2 = 4. The console returns the answer of the math problem on the next line.

### Task 3: Install packages

**Task** In the console, run the following code to install the Tidyverse:

```{r}
install.packages('tidyverse')
```

**Task** Add that

Then run the following code to install the rmarkdown library

```{r}
install.packages('rmarkdown')
```

## Chapter 6

### Task 1: Load libraries

**Task** Run the following code in the gray-colored codeblock below -- not in the console -- to load the tidyverse library. To run the code, click the little green play button (left facing arrow) at the top right of the codeblock. In Rmarkdown data notebooks, we write code inside of codeblocks, and explanatory text in the white area outside of it.

```{r}
# This is a comment inside of a codeblock. Comments doesn't run when we run the codeblock.
# Adding this command is a good way to start your class reference notebook.
library(tidyverse)
```

### Task 2: Load data

**Task** Load the Maryland WinRed contributions dataset by running the following codeblock.

```{r}
maryland_winred_contributions <- read_rds("pre_lab_01.rds")
```

Once it loads, look at the environment window at right in RStudio for maryland_winred_contributions. It's next to a small blue circle with a white arrow in it. To the right of that, look at how many "obs" the dataset has ("obs" is short for observation, which is another way of saying rows or records). Look at how many variables it has (another way of saying columns or fields). Write the number of observations and variables below.

**Answer**
Answer: There are 131,395 obs or observations in the data set which means there are 131,395 rows or records. The data set also contains 24 variables/fields. 

### Task 3: Glimpse data

**Task** Run the following codeblock to use glimpse() to get a sense of the data, some of the values in each field/variable/column, and the data types of each field/variable/column.

```{r}
glimpse(maryland_winred_contributions)
```

In the space below, write the data type for the "last_name" column and the "amount" columns (it's in light gray, offset by \< \>). Also offer your best guess as to what the "last_name" column and the "amount" column represent. **Answer**
Answer: The data type for the "last_name" column is a character which means it is specifying a character in a variable. The "amount" column is a dbl data type meaning it is a quantitative variable that contains decimal values. I think that the "last_name" column represents the last name of votes and the "amount" column represents how much money they have contributed to the Maryland WinRed foundation.

### Task 4: Examine data "head"

**Task** Run the following codeblock to use head() to examine the first six rows of data. Explore the data by clicking the black arrow on the right side of the output table, allowing you to examine additional columns.

```{r}
head(maryland_winred_contributions)
```

In the space below, write the first and last name from the _second_ row of the data. **Answer**
Answer: Peggy Mukai is the first and last name from the second row of the data. 

### Task 5: Explore data with interactive dataframe

**Task** look at the environment window at right in RStudio for maryland_winred_contributions It's next to a small blue circle with a white arrow in it. Click on the maryland_winred_contributions -- not the blue circle. It will open the dataframe as an interactive table. Scroll down to see additional rows. Scroll left to see additional columns. Click on a column name to sort by that column.

### Task 6: Group and count

**Task** Run the following codeblock:

```{r}
maryland_winred_contributions |>
  group_by(date) |>
  summarise(
    count_contribs = n()
  )
```

Describe what you think this code is doing below. What does the answer represent, compared to the results from `head()`?
**Answer**
Answer: This code is grouping the data by date and breaking it up into sets of 10 rows due to the count_contribs = n() function. Head gives us the first 6 lines of the data, grouping it into columns of six to make it more manageable to scroll through. Both blocks of code use different functions to group up the code in alternative, each of which may be specifically helpful for a user depending on what they are hoping to collect from the data. 

### Task 7: Adding arrange()

**Task** Run the following codeblock:

```{r}
maryland_winred_contributions |>
  group_by(date) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))
```

Compared to the previous codeblock, what has changed and how did that affect the result?

**Answer**
Answer: Using the group_by and summarise() functions allowed us to condense our data into groupings. In this block of code we have introduced the arrange function. The arrange function arranges our data and gives us meaning or purpose to tell R what we want to see first. This changed the data for the better in that it is much easier/cleaner for us to understand and read. 

### Task 8: Multiple grouping

**Task** Run the following codeblock.

```{r}
maryland_winred_contributions |>
  group_by(memo_text, amount) |>
  summarise(
    count_contribs = n()
  ) |>
  arrange(desc(count_contribs))
```

Compared to the previous codeblock, what has changed and how did that affect the result?

**Answer**
Answer: In this codeblock we are grouping by memo_text and amount. This affected the results in that it allows us to group by two things in combination and together. 

### Task 9

**Task** Run the following codeblock:

```{r}
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount)
  ) |>
  arrange(desc(total_amount))
```

Compared to the previous codeblock, what has changed and how did that affect the result? 

**Answer**
Answer: In task 9 we are adding total_amount = sum(amount). This change is noticeable in the labeling of the columns. In the previous codeblock the columns were labeled "memo_text", "amount", and "count_contribs". Now, with the alterations to the code block, the columns are labeled "memo_text", "count_contribs", and "total_amount". These changes affect the result because it is calling for the data to be arranged based on different variables. 

### Task 10: More descriptive statistics

**Task** Run the following codeblock to group the maryland_winred_contributions dataframe by the recipient for each contribution. Then do the following summarizations for each group:count the number of contributions, add up the value of all contributions, and calculate the mean and median contribution amounts. Then sort so the recipient with the most contributions is listed first.

```{r}
maryland_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount)
  ) |>
  arrange(desc(count_contribs))
```

Compared to the previous codeblock, what has changed and how did that affect the result? Glancing at the names, which of the top results from Maryland?

**Answer**
Answer: The differences between this codeblock and the previous code block are very apparent visually and functionally. In this codeblock we have access to the "mean_amount" and "median_amount" in addition to the "sum_amount" which was included in the former codeblock. We added median(amount) and mean(amount) and by adding these functions are expanding what the data set tells us.
### Task 11

**Task** Run the following codeblock to group the maryland_winred_contributions dataframe by the city for each contribution. Then do the following summarizations for each group:count the number of contributions, add up the value of all contributions, calculate the mean and median contributions amounts and calculate the smallest and largest contributions Then sort so the city with the largest total contribution amount is listed first.

```{r}
maryland_winred_contributions |>
  group_by(city) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount),
    min_amount = min(amount),
    max_amount = max(amount)
  ) |>
  arrange(desc(total_amount))
```

Compared to the previous codeblock, what has changed and how did that affect the result? How is the data sorted? **Answer**
Answer: In this codeblock, instead of grouping the data by "memo_text" like we did previously, we are grouping the data by "city". The other change that we are making here is that we are adding the functions, min(amount) and max(amount) to tell break down and simplify our data even more for easy access to important numbers. This change is also reflected by two new columns named, "min_amount" and "max_amount". 

