---
title: "lab_04"
author: "Adelia McGuire"
date: "2024-09-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(janitor)
library(lubridate)

```

For this lab, we want to investigate spending by the two leading Senate candidates in Maryland during this election, Angela Alsobrooks and Larry Hogan. Our goal is to try and standardize the purpose of the expenditures so that we can compare the two campaigns. We have the data, but it's messy - purpose descriptions are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of campaign expenditures located in the data folder called "md_senate_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
cleaned_data <- read_csv("~/Downloads/GitHub JOUR472/GitHub Folder/data_journalism_2024_fall/labs/lab_04/data/md_senate_expenses.csv")

cleaned_md_senate_expenses <- cleaned_data |>
  mutate(across(where(is.character), ~na_if(., "")), date = ymd(date))

str(cleaned_md_senate_expenses)

head(cleaned_md_senate_expenses)
```

## Answer questions

**Q1.** You've been assigned to compare the spending priorities of the two campaigns, but they describe their spending differently using the `purpose` column. Using OpenRefine, create a project using the original CSV file, make a copy of the `purpose` column called `purpose_clean` and then standardize it, focusing on the purposes that appear most often or have the largest total amounts. You can transform the data as you want - making all records the same case, for example, to help. The goal is to be able to group together as many similar purpose values, so you can choose to call all transportation-related expenses the same thing.

How do you know when you're done? It's a judgment call, but you should be close to 100 unique values for `purpose_clean`, and maybe even below that number.

Then, when you are done standardizing `purpose_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.

Then, using that `purpose_clean` column, write code to find the largest amounts of spending on categories by both campaigns.

**A1.** After using OpenRefine to clean the data within the "purpose" column I was able to merge/group together similar purpose values made by each of the two campaigns. I found that the Alsobrooks campaign's top spending purpose was Online/Media/Radio Advertising where a grand total of 4122402.70 was spent. Though Hogans campaign spent substantially less at just 2315936.48, Online/Media/Radio Advertising is also where his campaign spent the most.  

```{r}
cleaned_purpose <- read_csv("~/Downloads/GitHub JOUR472/GitHub Folder/data_journalism_2024_fall/labs/lab_04/data/cleaned-purpose-md-senate-expenses-csv.csv")

spending_summary <- cleaned_purpose |>
  group_by(cleaned_purpose,candidate, organization_name) |>
  summarise(total_spent = sum(amount, na.rm = TRUE), .groups = 'drop') |>
  arrange(desc(total_spent))

print(spending_summary)
```
```{r}
top_spenders <- spending_summary |>
  group_by(cleaned_purpose) |>
  summarise(total_spent = sum(total_spent), .groups = 'drop') |>
  arrange(desc(total_spent))

print(top_spenders)
```

**Q2.** Let's do some research on the top purposes from A1. Feel free to search the Web to understand those terms better if need be, and then write a description comparing and contrasting the two campaigns' largest categories of spending.

**A2.** Through analyzing the top few characters within the cleaned_purpose column, I can identify several words/terms that I am not very familiar with. Through my research on the web I found that "List Acquisition", which is positioned fourth in receiving the highest amount of expenses, is the process of finding and delivering lists of potential customers and/or donors for marketing campaigns. List acquisition is important in that it can help  create lists that will bring you success in direct marketing campaigns. Another term within the purpose column that I did not know much about was "Consulting Services" which I discovered through web searches is a broad term encompassing charges for various professional services that compensate consultants for their time, knowledge, expertise, etc. Both of the campaigns highest category for spending is in Online/Media/Radio Advertising. Advertising of any sort is a huge part of campaign success.  I am not surprised to see that this area is responsible for the highest expenditures for both Alsobrooks and Hogan as advertising yourself, your values, and your campaign is critical to success. 

**Q3.** Pick one of the top categories for each campaign and write code that shows how that spending has changed over time, if at all. You'll need to use lubridate for this. Write a couple of sentences describing that pattern.

**A3.**

```{r}
spending_timeline <- cleaned_purpose |>
  filter(cleaned_purpose == "Online/Media/Radio Advertising") |>
  mutate(date = ymd(date)) |> 
  group_by(date, candidate) |>
  summarise(total_spent = sum(amount, na.rm = TRUE), .groups = 'drop') |>
  arrange(date)

print(spending_timeline)
```

**Q4.** Which committee has the larger payroll or salary spending so far? How many unique recipients of payroll/staff spending are there for each candidate?

**A4.** The Alsobrooks committee has the larger payroll/salary spending so far, coming in at 8911505 compared to the Hogan committee with 4340010. While the payroll/salary spending of Alsobrooks and Hogan differ quite greatly, the amount of unique recipients of payroll/staff spending for each committee is very similar. Alsobrooks has 200 unique recipients and Hogan has 192, putting them at an 8 recipient differential. 

```{r}
payroll_salary_summary <- cleaned_purpose_md_senate_expenses_csv |>
  group_by(candidate) |>
  summarise(total_payroll = sum(amount, na.rm = TRUE), unique_recipients = n_distinct(organization_name), .groups = 'drop')

print(payroll_salary_summary)
```

**Q5** What's the most interesting story idea you've found in this data?

**A5** One potentially interesting story that this data holds pertains to the differing timelines of spending between the two candidates. Angela Alsobrooks made her first campaign-related expenditure on May 16, 2023, and continued to make 16 additional spending entries up until March 11, 2024. In contrast, Larry Hogan did not make any campaign expenditures until March 12, 2024. I feel that this expenditure timeline related discrepancy raises intriguing questions about their respective campaign strategies. Why did Alsobrooks begin spending so early in the race or rather what was Hogan's rationale for delaying his expenditures. Exploring these questions could uncover valuable insights into the effectiveness of their strategies. This data driven story idea highlights the differing approaches of the candidates and shows the significance of time in its influence over voter outreach. 
